### 2021-07-04 rad-icfp-2021/purerad.tex
minor

TODO:redo
In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
Admittedly, this transformation may be hard to read, and prove correct. 
In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).

\begin{example}
    TODO
\end{example}

We introduce several notations which are useful when defining the transformation for reverse-mode.

@@

@@

Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
and we need to initialise the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
which we initialise at $1$ to run the backpropagation. 
All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.

\begin{example}
    TODO
\end{example}

Following the insight from Section~\ref{subsec:insights}, 
we don't want to keep all the lambda abstractions as this is costly. 
Fortunately, the transformation is designed in such a way that we only have applications 
and we can use partial evaluation to do beta redudction.

@@

TODO: redo
Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
which garantees that this will not blow up the complexity.

@@

Though this transformation after the inlining above already has the right complexity, 
its purity allows us to take the most out of generic optimisations.  
A list of common optimisations that are useful for this language can be found in Figure~\ref{fig:optim}.

TODO: use the optimisations to be efficient on examples

>>>>

In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimize it further and compute the gradient.
Admittedly, this transformation may be hard to read, and it is not straightforward to show its correctness directly. 
In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).

\begin{example}
    The reverse-mode transformation of the terms from the introduction is

    \begin{tabular}{c r}
        &$\directD{\rho}{\Gamma}{Y}$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
        =& TODO \\
        \\
        $\directD{\rho}{\Gamma}{Y}$(!prod(A)!) \\
        =& \\
        % !dot(A,B)!
    \end{tabular}
\end{example}

The idea is that $\rho$ represents the return type of the derivative part, which should be !A1x...,xAn! 
if we want the whole gradient of a term !e! in context $\Gamma$=!x1:A1,...,xn:An!.
The subscript $\Gamma$ represents the current context, 
which is locally augmented for instance when differentiating a !let! rule.
For non-unary operations, we differentiate the arguents from left to right and add their derivatives to the current stack, 
which is modeled by the continuation $Y$. 
Importantly for performance, each continuation variable $Y$ is only used once.

We now introduce several notations which are useful when defining the transformation for reverse-mode.

@@

We have the following typing lemma for $\directD{\rho}{\Gamma}{Y}$.
\begin{lemma}[Typing $\directD{\rho}{\Gamma}{Y}$]
    If $\Gamma \vdash$ !e: A!, then $\Gamma$!,Y: !$\Gamma\to \rho \vdash \directD{\rho}{\Gamma}{Y}$!(e): !$\directD{\rho}{\Gamma}{Y}$!(A)!.
\end{lemma}

@@

TODO: redo
Given a term $\Gamma\vdash e : \reals$, we can compute its gradient $\grad_\Gamma e$ from a particular instance of 
$\directD{\rho}{\Gamma}{Y}(e)$. First, $\rho, Y$ specifies if we want to compute the whole gradient regarding the variables from $\Gamma$ or a subset of it.
For a subset $\rho\subset \Gamma$, one chooses $Y$ to be the projection function sending a variable 
$x_i:G$ of $\Gamma$ to $x_i$ if it belongs to $\rho$ and to $0_G$ otherwise.
In particular, we take $Y=Id_\Gamma$ to compute the whole gradient.
Next, the gradient will be given by the second part of the pair $\directD{\rho}{\Gamma}{Y}(e)$, 
and we need to initialize the tangent variables. All of them are set to $0$ except the one corresponding to the output value of !e!, 
which we initialize at $1$ to run the backpropagation. 
All in all, we compute the gradient via $\pi_2\directD{\rho}{\Gamma}{Id_\Gamma}(e)(0_\Gamma,1)$.

\begin{example}
    The gradient of the terms from the introduction is

    \begin{tabular}{c r}
        & $\grad_\Gamma$(!let w1 = x1 * x2 in let w2 = w1 * x1 in w2!) \\
        =& TODO \\
        \\
        $\grad_\Gamma$(!prod(A)!) \\
        =& \\
        % !dot(A,B)!
    \end{tabular}
\end{example}

Following the insight from Section~\ref{subsec:insights}, 
we don't want to keep all the lambda abstractions as this is costly. 
Fortunately, the transformation is designed in such a way that we only have applications 
and we can use partial evaluation to do beta-reduction.

@@

Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
which garantees that this will not blow up the complexity.

@@

Though this transformation after the inlining above already has the right complexity, 
its purity allows us to make the most out of generic optimizations.  
A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.

TODO: use the optimizations to be efficient on examples
